---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# maxRgain

maxRgain implements an Integer Programming-based method for optimising genetic gain in polyclonal selection, where the goal is to select a group of genotypes that jointly meet multi-trait selection criteria. The method uses predictors of genotypic effects obtained from the fitting of mixed models. Its application is demonstrated with grapevine data, but is applicable to other species and breeding contexts.

## Key Features

* Integer programming approach to group selection under multiple trait constraints

* Selection based on predictors of genotypic effects from mixed models

* Examples with real data from field trials of grapevine 

* The method can be generalized to a variety of crops and breeding programs

* Includes example dataset and reproducible workflows

## Installation

To install the development version of maxRgain from [GitHub](https://github.com/):

``` {r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("soniasurgy/maxRgain")
```
To install from CRAN:

```{r, eval = FALSE}
install.packages("maxRgain")
```

## Example

This is a basic example which shows how to solve a common problem:

```{r example}
library(maxRgain)
data("Gouveio")
mydmg <- data.frame(
  lhs = c("yd", "pa", "ta"),
  rel = c(">=", ">=", ">="),
  rhs = c(20, 3, 3),
  stringsAsFactors = FALSE
  )

polyclonal(
  traits = c("yd", "pa", "ta"),
  clmin = 7,
  clmax = 12,
  dmg = mydmg,
  meanvec = c(yd = 3.517, pa = 12.760, ta = 4.495),
  data = Gouveio
  )

```
For detailed guidance, see the package vignette:

```{r, eval = FALSE}
vignette("maxRgain", package = "maxRgain")
```
## Citation

The underlying method and this package can be cited as follows:

**Method:**  
Surgy, S., Cadima, J. & Gonçalves, E. Integer programming as a powerful tool for polyclonal selection in ancient grapevine varieties. Theor Appl Genet 138, 122 (2025). https://doi.org/10.1007/s00122-025-04885-0

**Package (In preparation):**  
Surgy, S., Cadima, J. & Gonçalves, E. maxRgain - A package to maximize genetic gains of polyclonal selection. Manuscript in preparation.

## License

This package is released under the GPL-3 License.

## Contact and Bug reports

For suggestions, and bug reports use the link:

https://github.com/soniasurgy/maxRgain/issues



