---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# maxRgain

<!-- badges: start -->
<!-- badges: end -->

maxRgain implements an integer programming-based method for optimizing genetic gain in polyclonal selection, where the goal is to select a group of genotypes that collectively meet multi-trait selection criteria. The method relies on genetic effect predictors obtained from mixed models and is demonstrated using grapevine breeding data, although it is broadly applicable to other species and breeding contexts.
This package offers a ready-to-use implementation of the method proposed by Surgy et al. (2025), allowing breeders to perform group-based multi-trait selection in a consistent and computationally efficient manner. It supports balanced genetic improvement across traits and fosters transparency in breeding decision-making.

## Key Features

* Integer programming approach to group selection under multiple trait constraints

* Selection based on genetic effect predictors from mixed models

* Examples with real grapevine breeding data

* The method can be generalized to a variety of crops and breeding programs

* Includes example dataset and reproducible workflows

## Installation

To install the development version of maxRgain from [GitHub](https://github.com/):

``` {r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("soniasurgy/maxRgain")
```
To install from CRAN:

```{r, eval = FALSE}
install.packages("maxRgain")
```

## Example

This is a basic example which shows how to solve a common problem:

```{r example}
library(maxRgain)
data("Gouveio")
mydmg <- data.frame(
  lhs = c("yd", "pa", "ta"),
  rel = c(">=", ">=", ">="),
  rhs = c(20, 3, 3),
  stringsAsFactors = FALSE
  )

polyclonal(
  traits = c("yd", "pa", "ta"),
  clmin = 7,
  clmax = 12,
  dmg = mydmg,
  meanvec = c(yd = 3.517, pa = 12.760, ta = 4.495),
  data = Gouveio
  )

```
For detailed guidance, see the package vignette:

```{r, eval = FALSE}
vignette("Introduction to maxRgain", package = "maxRgain")
```
## Citation

The method and the software can be cited as follows:

**Method:**  
Surgy, S., Cadima, J. & Gonçalves, E. Integer programming as a powerful tool for polyclonal selection in ancient grapevine varieties. Theor Appl Genet 138, 122 (2025). https://doi.org/10.1007/s00122-025-04885-0

**Software (In preparation):**  
Surgy, S., Cadima, J. & Gonçalves, E. maxRgain - A package to perform polyclonal selection. Manuscript in preparation.

## License

This package is released under the GPL-3 License.

## Contact and Bug reports

For suggestions, and bug reports use the link:

https://github.com/soniasurgy/maxRgain/issues



