---
title: "Introduction to maxRgain"
author: "Sónia Surgy, Jorge Cadima, Elsa Gonçalves"
output: 
  html_document:
    toc: true
    toc_depth: 2

vignette: >
  %\VignetteIndexEntry{Introduction to maxRgain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(maxRgain)
```
# Introduction

Polyclonal selection is a breeding strategy developed in the context of ancient grapevine varieties, aimed at selecting a group of distinct genotypes that collectively satisfy predefined criteria for desirable quantitative traits. This approach relies on the predictors of the genetic effects obtained from the fitting of a mixed model. Rather than focusing on individual genotypes, polyclonal selection considers the group as the unit of selection, with the goal of maximising the overall predicted genetic gain across multiple traits of interest. The identification of such a superior group constitutes an optimisation problem, in which the objective is to select a subset of genotypes that jointly maximise the expected genetic value, subject to constraints which satisfy agronomic and enological interests.

Most existing methods for genetic selection based on several traits, such as selection indices, are designed to maximize the genetic gain of selection of a genotype. When the objective is to select a group of genotypes, rather than individuals independently, the conventional approach has been to simply choose the top-ranking genotypes. However, this strategy does not ensure that the selected group, as a whole, satisfies predefined breeding objectives when multiple traits are considered simultaneously.

This package implements a novel Integer Programming-based method developed by Surgy et al. (2025), designed specifically for group-based selection under multiple trait criteria. The method enables the identification of combinations of genotypes  that collectively meet desired selection goals, offering an alternative to traditional individual-based approaches. While the methodology is demonstrated using data from grapevine breeding trials, it is broadly applicable across different species and breeding programs, especially in scenarios where group performance is more relevant than individual merit alone.

For a complete description of the methodology, please refer to:

Surgy, S., Cadima, J. & Gonçalves, E. Integer programming as a powerful tool for polyclonal selection in ancient grapevine varieties. Theor Appl Genet 138, 122 (2025). https://doi.org/10.1007/s00122-025-04885-0

Each function in the package includes an example that reproduces scenarios described in the original publication, thereby facilitating both practical understanding and direct application of the method. This document provides a more detailed explanation of the outputs generated by the various functions and offers guidance on how to interpret them effectively.

# Data included

The package includes a real dataset—Gouveio—to support reproducibility of the examples and to allow users to explore additional scenarios. This dataset contains Empirical Best Linear Unbiased Predictors (EBLUPs) of genotypic effects, obtained from the fitting of a mixed model fitted to phenotypic data from a selection field trial involving 150 genotypes of the grapevine variety Gouveio established according to a resolvable row-column design.


```{r}
head(Gouveio)

```

The traits evaluated were:
* Yield (yd)
* Potential Alcohol (pa)
* Total Acidity (ta)
* pH
* Berry Weight (bw)

The selection criteria are defined by the breede. In this example will be: increasing yield, potential alcohol, and total acidity, while decreasing pH and berry weight.


# Common arguments

Below is a table describing the common arguments used across the functions in this package:

| Argument     | Description    | Used in functions  |
|--------------|------------------------------|-------------|
| `traits` | A vector with the traits to be optimized, i.e., those included in the objective function| All |
| `ref`        | Name of the reference column (e.g., genotype ID). Defaults to the first column. | All |
| `clmin` |  Minimum group size. |`polyclonal()` |
| `clmax` |  Maximum group size. If omitted, equal to `clmin` |`polyclonal()` |
| `dmg`        |  A *data.frame* with three columns defining constraints: trait names, constraint signs (`">="`, `"<="`, `"=="`), and *right-hand side* values. |`polyclonal()` |
| `constraints`| Vector with traits to which constraints apply. If omitted, all except `ref` are used. | `BaseSituation()`, `Rmaxa()` |
| `meanvec`    |  Vector of overall means per trait; if omitted, data are assumed to be already normalized by the mean. |All   |
| `criteria`   |  Vector with the criterion for each trait: 1 to increase, -1 to decrease. If omitted, all traits are assumed to be increasing. |All   |
| `data`       | A *data.frame* with the predictors of genotypic effects for the several traits. Rows are genotypes; columns are traits.   | All |


# Important notes

* If the gain for a group of a certain size is not displayed, it means that it was not possible to satisfy all the constraints for that group size. It is possible that constraints are not feasible for some group sizes but are feasible for others.

* The genotypes selected in a group of a given size may differ substantially from those selected in a group of another size. Changing any condition — including the group size — results in a new optimisation problem.

* The *data.frame* used for the `dmg` argument can have any column names, but the order must be respected: first the trait, then the signal (e.g., ">="), and finally the *right-hand side* value, that is, the value for the minimum desired gain in percentage of the overall mean of the trait.

* If a decrease in a trait is admissible, the corresponding value must be indicated as a negative number in the `dmg`.

* If no group satisfies the specified conditions, the message 'No possible solution!' is returned.

# Functions 

## `polyclonal()` 

This is the main function of this package. 

```{r, eval = FALSE}
polyclonal(traits, ref = NULL, clmin = 7, clmax,  dmg = NULL, meanvec = NULL, criteria = NULL, data)
```

In the example for the `polyclonal()` function, the objective is to maximize the genetic gain of groups with sizes ranging from 7 to 20 genotypes, while enforcing minimum desired gains (expressed as percentages of the overall mean) for the five traits: 20% for yield (yd), 3% for potential alcohol (pa), 3% for total acidity (ta), 1% for pH, and 2% for berry weight (bw).

```{r}
# Vector with the means
mymeanvec <- c(yd = 3.517, pa = 12.760, ta = 4.495, ph = 3.927, bw = 1.653)

# Vector with the traits to be maximized
mytraits <- c("yd", "pa",  "ta", "ph", "bw")

# Vector with the criteria
mycriteria <- c(yd = 1, pa = 1, ta = 1, ph = -1, bw = -1)

# Vector with the column with the names given to the genotypes
myref = "Clone"
```
If `meanvec` is omitted, it is assumed that all trait values have already been standardised by dividing by their respective overall means. If `criteria` is omitted, the default assumption is that an increase is desired for all traits. If `ref` is omitted, the first column in the dataset is assumed to contain the names of the genotypes."

In `polyclonal()` function the constraints are indicated by argument `dmg` (desired minimum gains). This argument consists in a *data.frame* composed by three columns: the first with the traits of the constraints; the second with the signal of the constraint (">=", "<=" or "=="); and the third with the values for the *right-hand side* of the constraints. The column names can be arbitrary, but their order must be preserved.
```{r}
# data.frame with the desired minimum gains
mydmg <- data.frame(
  lhs = c("yd", "pa", "ta", "ph", "bw"),
  rel = c(">=", ">=", ">=", ">=", ">="),
  rhs = c(20, 3, 3, 1, 2),
  stringsAsFactors = FALSE
  )
```
By calling `mydmg` it is easy to verify whether the specified values have been correctly applied to the desired constraints.
```{r}
mydmg

```
Now, the `polyclonal` function is called to perform the selection based on the specified constraints:
```{r}
# Using polyclonal() function
with_dmg <- polyclonal(
  traits = mytraits,
  clmin = 7,
  clmax = 20,
  dmg = mydmg,
  meanvec = mymeanvec,
  criteria = mycriteria,
  data = Gouveio
  )
```

Groups ranging from 7 to 20 genotypes were requested, as these cover the most appropriate group sizes for practical applications of the polyclonal methodology for grapevine selection. However, other group sizes can also be specified. In fact, it is possible to request a single group size by assigning the same value to both `clmin` and `clmax` or omitting `clmax`. Analysing the output reveals how the imposed constraints influence the selection results.
```{r}
# Object gain
with_dmg$gain
```

The output displays the genetic gain achieved for each trait across the different group sizes. For example, in the group composed of 10 genotypes, the gains obtained were approximately 21.4%, 3.0%, 3.3%, 1.1% and 2.3% for yield (yd), potential alcohol (pa), total acidity (ta), pH, and berry weight (bw), respectively. All values are expressed as percentages of the overall trait mean.  Positive values indicate gains in accordance with the selection criteria.

In this case, although group sizes between 7 and 20 genotypes were requested, the object `gain` only provides results for groups of 7, 8, and 10 genotypes. This indicates that the desired minimum gains could not be achieved for other group sizes within the specified range. 

```{r}
# Object selected
with_dmg$selected
```
Examining the lists of genotypes composing each group reveals that a genotype present in a group of one size may be absent from groups of different sizes. For example, genotype GV060 is not included in the group of 7 genotypes, appears in the group of 8 genotypes, and is then excluded from the group of 10 genotypes.


## `BaseSituation()` 


```{r, eval = FALSE}
BaseSituation(traits, ref = NULL, constraints = NULL, meanvec = NULL, criteria = NULL, data)
```

The base situation relies on the use of constraints to prevent losses in all traits of interest. In this case, one constraint is defined for each trait included in the constraint set. The traits subject to constraints may or may not coincide with those targeted for maximization. In the base situation, all constraints associated with the traits must be greater than or equal to zero, ensuring no negative gain in any of them.

The package includes a function, `BaseSituation()`, which automatically computes this configuration for group sizes ranging from 7 to 20 genotypes.


```{r}
# mymeanvec, mytraits and mycriteria, remains the same

# Vector with the means
mymeanvec <- c(yd = 3.517, pa = 12.760, ta = 4.495, ph = 3.927, bw = 1.653)

# Vector with the traits
mytraits <- c("yd", "pa","ta", "ph", "bw")

# Vector with the criteria
mycriteria <- c(yd = 1, pa = 1, ta = 1, ph = -1, bw = -1)

# Vector with the traits used for the constraints
restrict <- c("yd", "pa", "ta", "ph", "bw")
```
For the `BaseSituation()` function, if the `constraints` argument is omitted, all columns in the dataset are considered as constrained traits, except the one specified by `ref`. If `ref` is also omitted, the first column is assumed to correspond to the reference column.
```{r}
# Using the BaseSituation() function:
base_sit <- BaseSituation(
  traits = mytraits,
  constraints = restrict,
  meanvec = mymeanvec,
  criteria = mycriteria,
  data = Gouveio
  )

# Object gain
base_sit$gain

```
Comparing the two scenarios highlights the impact of the constraints: when using the BaseSituation() function, no losses are observed in any of the traits. Additionally, the comparison reveals differences in the composition of the selected groups, as the genotypes chosen under the constrained scenario often differ from those selected without constraints.
```{r}
# Object selected
base_sit$selected
```
As an example, it is possible to analyse the groups of 7 and 8 genotypes. In the group of 8 genotypes, there are not just one but two genotypes that differ from the group of 7: genotypes GV089 and GV137. On the other hand, genotype GV132, which is included in the group of 7 genotypes, is not present in the group of 8. This illustrates how the inclusion of constraints can lead to changes in group composition, rather than simply adding the next best-ranked individual.


## `Rmaxp()` 

```{r, eval = FALSE}
Rmaxp(traits, ref = NULL, meanvec = NULL, criteria = NULL, data)
```



This function computes the maximum possible gain for group sizes ranging from 7 to 20 genotypes. The maximum possible gain corresponds to the value obtained by selecting the top genotypes for each trait individually, without applying any constraints. The function allows the user to request the gain for one or multiple traits simultaneously. The output displays one column per trait; however, the values in different columns are not linked to the same solution and should be interpreted independently. 
```{r}
# Vector with the means
mymeanvec <- c(yd = 3.517, pa = 12.760)

# Vector with the target traits
mytraits <- c("yd", "pa")
```
The vector specifying the selection criteria can be omitted, as the default value of 1 is used for all traits. With this function, no constraints are applied.

```{r}
# Using Rmaxp()
max_pos_gain <- Rmaxp(
  traits = mytraits,
  meanvec = mymeanvec,
  data = Gouveio
  )

# Object gain
max_pos_gain$gain
```
As in previous outputs, one column per trait is shown. However, in this case, the gain reported in each column refers to an independent solution. For example, the group of 20 genotypes achieving a gain of 39.5% for yd is not composed of the same genotypes as the group achieving a gain of 5.9% for pa. This can be verified by examining the output stored in the selected object.

```{r}
# Object selected
max_pos_gain$selected
```

Since no constraints are applied in this scenario, the output represents a ranking of genotypes for each trait. Increasing the group size simply adds the next genotype in the ranking. Thus, each column corresponds to the top genotypes for a given trait. The final column, Entry, indicates the smallest group size at which each genotype is first included. For example, genotype GV137 appears for the first time in yd when the group size reaches 10 genotypes; when the group increases to 11, GV137 remains in the group and genotype GV079 is added.

## `Rmaxa()` 

```{r , eval=FALSE}
Rmaxa(traits, ref = NULL, constraints = NULL, meanvec = NULL, criteria = NULL, data)

```


This function computes the maximum admissible gain for group sizes ranging from 7 to 20 genotypes. The maximum admissible gain corresponds to the value obtained by maximising one trait while avoiding losses in all others, enforced through constraints—predefined in this case as *>= 0*. As with `Rmaxp()`, the user may request gains for one or multiple traits simultaneously. The output displays one column per trait; however, the values in different columns correspond to independent solutions and should be interpreted separately. This distinguishes the output of the gain object from this function from that of the `BaseSituation()` function, where all traits are optimized together.

```{r}
# Vector with the means
mymeanvec <- c(yd = 3.517, pa = 12.760, ta = 4.495, ph = 3.927, bw = 1.653)

# Vector with the target traits
mytraits <- c("yd", "pa")

# Vector with the criteria 
mycriteria <- c(yd = 1, pa = 1, ta = 1, ph = -1, bw = -1)
```
As with the `BaseSituation()` function, when defining the constraints it is only necessary to specify the traits to which the constraints apply, since all are assumed to be greater than or equal to zero. If this vector is omitted, all columns in the dataset are considered as constrained traits, except the one indicated in `ref` or the first column if ref is omitted. The `criteria` as well as the `meanvec` arguments must indicate the criteria for all the traits involved in both maximization and the constraints.
```{r}
# Using Rmaxa()
max_adm_gain <- Rmaxa(
  traits = mytraits,
  meanvec = mymeanvec,
  data = Gouveio
  )

# Object gain
max_adm_gain$gain
```
As in the `Rmaxp()` gain output, one column per trait is shown; however, the gain reported in each column corresponds to an independent solution. For example, the group of 20 genotypes achieving a gain of 26.2% for *yd* is not composed of the same genotypes as the group achieving a gain of 5.7% for *pa*. Furthermore, the gains obtained with `Rmaxa()` for any trait are always less than or equal to those obtained with `Rmaxp()`, since `Rmaxa()` incorporates constraints during optimisation. 

Regarding the `selected` object, due to the constraints applied in this function, the genotypes composing one group differ from those in other groups. Consequently, there is one selected object per trait, named as `selected_<trait>`.

```{r}
# Object selected
max_adm_gain$selected_yd
```


# Conclusion

This package offers a ready-to-use implementation of the method developed by Surgy et al. (2025), allowing breeders to efficiently and consistently perform group-based multi-trait selection. With included examples and data, it provides an accessible tool to support balanced genetic gain across multiple traits in breeding programs.








